Why do species get a thin slice of $\pi$? Revisiting Lewontin's Paradox of Variation

### 2021-05-25 manuscript/manuscript.tex
abstract updated

  Under neutral theory, genetic diversity in an equilibrium population is
  expected to increase with population size. However, observed levels of
  diversity across metazoans vary only two orders of magnitude, while census
  population sizes ($N_c$) are expected to vary over several. This unexpectedly
  narrow range of diversity is a longstanding enigma in evolutionary genetics
  known as Lewontin's Paradox of Variation (1974). Since Lewontin's
  observation, it has been argued that selection constrains diversity across
  species, yet tests of this hypothesis seem to fall short of explaining the
  orders-of-magnitude reduction in diversity observed in nature. In this work,
  I revisit Lewontin's Paradox and assess whether current models of linked
  selection are capable of constraining diversity to this extent. To quantify
  the discrepancy between pairwise diversity and census population sizes across
  species, I combine genetic data from 172 metazoan taxa with estimates of
  census sizes from geographic occurrence data and population densities
  estimated from body mass.  Next, I fit the relationship between
  previously-published estimates of genomic diversity and these approximate
  census sizes to quantify Lewontin's Paradox. To account for phylogenetic
  non-independence, I use phylogenetic comparative methods to investigate the
  diversity census size relationship, estimate phylogenetic signal, and explore
  how diversity changes along the phylogeny. I consider whether the reduction
  in diversity predicted by models of recurrent hitchhiking and background
  selection could explain the observed pattern of diversity across species.
  Since the impact of linked selection is mediated by recombination map length,
  I also investigate how map lengths vary with census sizes. I find species
  with large census sizes have shorter map lengths, leading these species to
  likely experience greater reductions in diversity due to linked selection.
  Even after using high estimates of the strength of sweeps and background
  selection, I find linked selection likely cannot explain the shortfall
  between predicted and observed diversity levels across metazoan species.
  Furthermore, the predicted diversity under linked selection does not fit the
  observed diversity--census-size relationship, implying that processes other
  than background selection and recurrent hitchhiking must be limiting
  diversity.

>>>>

 Neutral theory predicts that genetic diversity increases with population size,
 yet observed levels of diversity across metazoans vary only two orders of
 magnitude while population sizes vary over several. This unexpectedly narrow
 range of diversity is known as Lewontin's Paradox of Variation (1974). While
 some have suggested selection could constrain diversity, tests of this
 hypothesis seem to fall short. Here, I revisit Lewontin's Paradox and assess
 whether current models of linked selection are capable of reducing
 diversity to this extent. To quantify the discrepancy between pairwise
 diversity and census population sizes across species, I combine
 previously-published estimates of pairwise diversity from 172 metazoan taxa
 with estimates of census sizes. Using phylogenetic comparative methods, I show
 this relationship is significant accounting for phylogeny, but with high
 phylogenetic signal and evidence that some lineages experience shifts in the
 evolutionary rate of diversity deep in the past. Additionally, I find a
 negative relationship between recombination map length and census size,
 suggesting abundant species have less recombination and experience greater
 reductions in diversity due to linked selection. However, I show that even
 using strong selection parameter estimates, models of linked selection are
 unlikely to explain the observed relationship between diversity and census
 sizes across species.
